### 2016-11-21 dissertation/prelim_GF.tex
dissertation v2_finish prelims

{\bf Ring} is a broad algebraic concept. In this dissertation, this word is used to refer a special 
sort of ring -- {\bf commutative ring with unity}. Two common 
examples of such rings are the set of integers, $\mathbb{Z}$, and the set of 
rational numbers, $\mathbb{Q}$. Note that while both of these examples are
rings with an infinite number of elements, the number of elements in a ring 
can also be finite.

@@

The main concept of field theory is {\bf Field Extensions}. The idea behind a
field extension is to take a base field and construct a larger field which 
contains the base field as well as satisfies additional properties. For example,
the set of real numbers $\mathbb{R}$ forms a field; one common extension of 
$\mathbb{R}$ is the set of complex numbers $\mathbb{C}=\mathbb{R}(i)$. Every
element of $\mathbb{C}$ can be represented as $a+b\cdot i$ where $a,b \in \mathbb{R}$,
hence $\mathbb{C}$ is a two-dimensional extension of $\mathbb{R}$.

Like rings, fields can also contain either an infinite or a finite number of 
elements. 
In this dissertation we focus on finite fields, also known as Galois fields, and 
the construction of their field extensions.

\subsection{Galois Field}
Galois fields, also known as finite fields, find widespread applications in 
many areas of electrical engineering and computer science such as error-
correcting codes, elliptic curve cryptography, digital signal processing, 
testing of VLSI circuits, among others.
In this dissertation, we specifically focus on their application to 
Elliptic Curve Cryptography as Galois field arithmetic circuits.
This section describes the relevant Galois field concepts
\cite{galois_field:mceliece} \cite{ftheory:2006} \cite{ff:1997}
and hardware arithmetic designs over such fields \cite{mastro:1989} \cite{PT:1985} 
\cite{acar:1998} \cite{wu:2002} \cite{Knezevic:2008}.

@@

Fields in the form $\F_{p^{k}}$ are called Galois extension fields.
We are specifically interested in extension fields of type 
$\Fkk$, where $k > 1$. These are extensions of the binary
field $\F_2$.
\begin{Example}
Addition and multiplication operations over $\F_2$:
\begin{table}[!h]
	\centering
	\begin{tabular}{m{1cm}|l|ll|m{1cm}}
	\hhline{~---~}
	\multirow{3}{*}{} & $+$ & $0$ & $1$ & \multirow{3}{*}{} \\
	\hhline{~---~}
	& $0$ & $0$ & $1$ & \\
	& $1$ & $1$ & $0$ & \\
	\hhline{~---~}
	\multicolumn{5}{c}{}\\
	\multicolumn{5}{c}{Addition over $\F_2$}\\
	\end{tabular}
	\quad
	\begin{tabular}{m{1cm}|l|ll|m{1cm}}
	\hhline{~---~}
	\multirow{3}{*}{} & $\cdot$ & $0$ & $1$ & \multirow{3}{*}{} \\
	\hhline{~---~}
	& $0$ & $0$ & $0$ & \\
	& $1$ & $0$ & $1$ & \\
	\hhline{~---~}
	\multicolumn{5}{c}{}\\
	\multicolumn{5}{c}{Multiplication over $\F_2$}\\
	\end{tabular}
\end{table}

@@

Algebraic extensions of the binary field $\F_{2}$  
are generally termed as {\it binary extension fields} $\Fkk$.
Where elements in $\F_2$ can only represent $1$ bit, elements in $\Fkk$ 
represent a $k$-bit vector.
This allows them to be widely used in digital hardware applications.
In order to construct a Galois field of the form $\Fkk$, 
an {\bf irreducible polynomial} is required:
\begin{Definition}
A polynomial $P(x) \in \mathbb{F}_{2}\left[x\right]$ is {\bf irreducible} 
if $P(x)$ is non-constant with degree $k$ and cannot be 
factored into a product of polynomials of lower degree in $\mathbb{F}_2[x]$.
\end{Definition}

@@

To better understand this field extension, compare its similarities to another
common-place
field extension $\C$, the set of complex numbers. $\C$ is an extension of the field 
of real numbers $\R$ with an additional element $i=\sqrt{-1}$, which is an imaginary
root in $\R$.
Thus $i \notin \R$, rather $i \in \C$.
Every element $A \in \mathbb{C}$ can be represented as:
\begin{equation}\label{rep:polyC}
A=\sum_{j=0}^{1} (a_j \cdot i^j)=a_0+a_1\cdot i
\end{equation}
where $a_j \in \R$ are coefficients. Similarly, $\Fkk$ is an extension of $\F_2$ with 
an additional element $\alpha$, which is the ``imaginary root'' of an irreducible 
polynomial $P$ in $\F_2[x]$.

@@

\begin{Definition}
A {\bf primitive polynomial} $P(x)$ is a polynomial with coefficients in $\mathbb{F}_2$ 
which has a root $\alpha$ $\in$ $\mathbb{F}_{2^k}$
such that \{$0$, $1(=\alpha^{{2^k}-1})$, $\alpha$, $\alpha^2$, $\cdots$, $\alpha^{2^k-2}$\} is the set of 
all elements in $\mathbb{F}_{2^k}$, 
where $\alpha$ is a {\bf primitive element} of $\mathbb{F}_{2^k}$. 
\end{Definition}

A primitive polynomial is guaranteed to generate all distinct elements 
of a finite field $\mathbb{F}_{2^k}$ while an irreducible polynomial
has no such guarantee.
Often, there exists more than one irreducible polynomial of degree $k$.
In such cases, any degree $k$ irreducible polynomial can be 
used for field construction. For example, both $x^3+x+1$ and $x^3+x^2+1$ 
are irreducible in $\mathbb{F}_2$ and either one can be used
to construct $\mathbb{F}_{2^3}$. This is due to the following:

@@

A Galois field $\F_q$ can be fully contained within a larger field $\F_{q^k}$.
That is, $\F_q \subset \F_{q^k}$.
For example, Fig \ref{fig:contain2_4_16} shows the containment of the fields 
$\F_2 \subset \F_4 \subset \F_{16}$. It's easy to see that since $\F_4=\F_{2^2}$, it
contains $\F_2$. Likewise $\F_{16}=\F_{4^2}=\F_{2^4}$ contains $\F_4$ and $\F_2$.
The elements $\{0,1,\alpha,\dots,\alpha^{14}\}$
designate $\F_{16}$. Of these, $\{0,1,\alpha^5,\alpha^{10}\}$ create $\F_4$.
From these, only $\{0,1\}$ exist in $\F_2$.

@@

>>>>

{\bf Ring} is a broad algebraic concept. In this dissertation, this word is used to refer a special 
sort of ring -- {\bf commutative ring with unity}. Two common 
examples of such rings are the set of integers, $\mathbb{Z}$, and the set of 
rational numbers, $\mathbb{Q}$. While both of these examples are
rings with an infinite number of elements, the number of elements in a ring 
can also be finite, such as the ring of integers modulo $n$ ($\Z_n$).

@@

An important concept in field theory is {\bf field extension}. The idea behind a
field extension is to take a base field and construct a larger field which 
contains the base field as well as satisfies additional properties. For example,
the set of real numbers $\mathbb{R}$ forms a field; one extension of 
$\mathbb{R}$ is the set of complex numbers $\mathbb{C}=\mathbb{R}(i)$. Every
element of $\mathbb{C}$ can be represented as $a+b\cdot i$ where $a,b \in \mathbb{R}$,
hence $\mathbb{C}$ is a two-dimensional extension of $\mathbb{R}$.

Like rings, fields can also contain either an infinite or a finite number of 
elements. 
In this dissertation we focus on finite fields -- also known as Galois fields, 
the construction of their field extensions, and their applications on circuit verification and abstraction techniques.

\subsection{Finite Field}
Finite fields find widespread applications in 
many areas of electrical engineering and computer science such as error-
correcting codes, elliptic curve cryptography, digital signal processing, 
testing of VLSI circuits, among others.
In this dissertation, we specifically focus on their application to 
the FSM traversal of sequential Galois field circuits as well as abstraction refinement based on UNSAT core extraction.
This section describes the relevant Galois field concepts
\cite{galois_field:mceliece} \cite{ftheory:2006} \cite{ff:1997}
and hardware arithmetic designs over such fields \cite{mastro:1989} \cite{PT:1985} 
\cite{acar:1998} \cite{wu:2002} \cite{Knezevic:2008}.

@@

Fields of the form $\F_{p^{k}}$ are called Galois extension fields.
We are specifically interested in extension fields of type 
$\Fkk$, where $k > 1$. These are extensions of the binary
field $\F_2$.
\begin{Example}
Addition and multiplication operations over $\F_2$:
\begin{table}[!h]
	\centering
	\begin{tabular}{m{1cm}|l|ll|m{1cm}}
	\hhline{~---~}
	\multirow{3}{*}{} & $+$ & $0$ & $1$ & \multirow{3}{*}{} \\
	\hhline{~---~}
	& $0$ & $0$ & $1$ & \\
	& $1$ & $1$ & $0$ & \\
	\hhline{~---~}
	\multicolumn{5}{c}{}\\
	\multicolumn{5}{c}{Addition over $\F_2$}\\
	\end{tabular}
	\quad
	\begin{tabular}{m{1cm}|l|ll|m{1cm}}
	\hhline{~---~}
	\multirow{3}{*}{} & $\cdot$ & $0$ & $1$ & \multirow{3}{*}{} \\
	\hhline{~---~}
	& $0$ & $0$ & $0$ & \\
	& $1$ & $0$ & $1$ & \\
	\hhline{~---~}
	\multicolumn{5}{c}{}\\
	\multicolumn{5}{c}{Multiplication over $\F_2$}\\
	\end{tabular}
\end{table}

@@

Algebraic extensions of the binary field $\F_{2}$  
are generally termed as {\it binary extension fields} $\Fkk$.
Where elements in $\F_2$ can only represent $1$ bit, elements in $\Fkk$ 
represent a $k$-bit vector.
This allows them to be widely used in digital hardware applications.
In order to construct a Galois field of the form $\Fkk$, 
an {\bf irreducible polynomial} is required:
\begin{Definition}
A polynomial $P(x) \in \mathbb{F}_{2}\left[x\right]$, i.e. the set of all 
polynomials in $x$ with coefficients in $\F_2$, is {\bf irreducible} 
if $P(x)$ is non-constant with degree $k$ and cannot be 
factored into a product of polynomials of lower degree in $\mathbb{F}_2[x]$.
\end{Definition}

@@

To better understand this field extension, compare its similarities to another
common-place
field extension $\C$, the set of complex numbers. $\C$ is an extension of the field 
of real numbers $\R$ with an additional element $i=\sqrt{-1}$, which is an imaginary
root in the algebraic closure of $\R$ -- the closure is known as the field of complex numbers $\C$.
Thus $i \notin \R$, rather $i \in \C$.
Every element $A \in \mathbb{C}$ can be represented as:
\begin{equation}\label{rep:polyC}
A=\sum_{j=0}^{1} (a_j \cdot i^j)=a_0+a_1\cdot i
\end{equation}
where $a_j \in \R$ are coefficients. Similarly, $\Fkk$ is an extension of $\F_2$ with 
an additional element $\alpha$, which is the ``imaginary root'' of an irreducible 
polynomial $P$ in $\F_2[x]$.

@@

\begin{Definition}
A {\bf primitive polynomial} $P(x)$ is a polynomial with coefficients in $\mathbb{F}_2$ 
which has a root $\alpha$ $\in$ $\mathbb{F}_{2^k}$
such that \{$0$, $1(=\alpha^{{2^k}-1})$, $\alpha$, $\alpha^2$, $\cdots$, $\alpha^{2^k-2}$\} is the set of 
all elements in $\mathbb{F}_{2^k}$.
Here $\alpha$ is called a {\bf primitive element} of $\mathbb{F}_{2^k}$. 
\end{Definition}

A primitive polynomial is guaranteed to generate all distinct elements 
of a finite field $\mathbb{F}_{2^k}$ while an arbitrary irreducible polynomial
has no such guarantee.
Often, there exists more than one irreducible polynomial of degree $k$.
In such cases, any degree $k$ irreducible polynomial can be 
used for field construction. For example, both $x^3+x+1$ and $x^3+x^2+1$ 
are irreducible in $\mathbb{F}_2$ and either one can be used
to construct $\mathbb{F}_{2^3}$. This is due to the following:

@@

A Galois field $\F_q$ can be fully contained within a larger field $\F_{q^k}$.
That is, $\F_q \subset \F_{q^k}$.
For example, the containment relation of the fields 
$\F_2 \subset \Fkk$ is usually used to represent bit-level Boolean variables as field elements
in larger finite field which allows projection of $k$-bit word-level variables. 
Concretely, $\F_{16}=\F_{4^2}=\F_{2^4}$ contains $\F_4$ and $\F_2$.
The elements $\{0,1,\alpha,\dots,\alpha^{14}\}$
designate $\F_{16}$. Of these, $\{0,1,\alpha^5,\alpha^{10}\}$ create $\F_4$.
From these, only $\{0,1\}$ exist in $\F_2$.

@@

\section{Normal Basis Multiplier over Galois Field}
From algebraic view, a field is a space, and field elements are points in the space. Those elements can be 
represented with unique coordinates, which requires the pre-definition of a basis vector. In this
section, we discuss a special basis called normal basis, as well as the advantages adopting it in 
GF operations esp. multiplication.
\subsection{Normal Basis}
Given a Galois field (GF) $\Fkk$ is a finite field with  $2^k$ elements and characteristic equals to 2.
Its elements can be written in polynomials of $\alpha$, when there is an irreducible polynomial $p(\alpha)$
defined.

If we use a basis $\{1,\alpha,\alpha^2,\alpha^3,\dots,\alpha^{k-1}\}$, we can easily transform polynomial representations
to binary bit-vector representations by recording the coefficients. For example,

\begin{table}[H]
\centering
\caption{Bit-vector, Exponential and Polynomial representation of
elements in  ${\mathbb{F}}_{2^4} = {\mathbb{F}}_2[x]
\pmod{x^4+x^3+1}$}
\begin{tabular}{|c|c||c|c|} 
\hline
$a_3a_2a_1a_0$ & Polynomial     &$a_3a_2a_1a_0$ & Polynomial  \\
\hline
$0000$        & $0$           & $1000$  &$\alpha^3$\\
\hline
$0001$        & $1$           & $1001$  & $\alpha^3 + 1$\\
\hline
$0010$        &  $\alpha$       & $1010$ & $\alpha^3 + \alpha$  \\
\hline
$0011$        &  $\alpha + 1$   & $1011$ &  $\alpha^3+\alpha+1$\\
\hline
$0100$        &  $\alpha^2$     &  $1100$ &  $\alpha^3 + \alpha^2$\\
\hline
$0101$        & $\alpha^2 + 1$ & $1101$  & $\alpha^3+\alpha^2+1$\\
\hline
$0110$        &  $\alpha^2 + \alpha$ & $1110$ &  $\alpha^3+\alpha^2+\alpha$\\
\hline
$0111$        & $\alpha^2+\alpha+1$ & $1111$ & $\alpha^3+\alpha^2+\alpha+1$\\
\hline
\end{tabular}
\label{table:booltogalois}  
\end{table}

Basis $\{1,\alpha,\alpha^2,\alpha^3,\dots,\alpha^{k-1}\}$ is called {\bf standard basis} (StdB), which results in
a straightforward representation for elements, and operations of elements such as addition and subtraction.
The addition/subtraction of GF elements in StdB follows the rules of polynomial addition/subtraction
where coefficients belong to $\mathbb F_2$. In other words, using the definition of {\it exclusive or} (XOR) in
Boolean algebra, element $A$ add/subtract by element $B$ in StdB is defined as
\begin{align}\label{eqn:StdB}
A+B = A-B &= (a_0,a_1,\dots,a_{k-1})_{StdB} \xor (b_0,b_1,\dots,b_{k-1})_{StdB} \nonumber\\
&=(a_0\oplus b_0, a_1\oplus b_1,\dots,a_{k-1}\oplus b_{k-1})_{StdB} 
\end{align}

\subsection{Multiplication using Normal Basis}
Besides addition/subtraction, multiplication is also very common in arithmetic circuit design.
The multiplication of GF elements in $\Fkk$ in StdB follows the rule of polynomial multiplication.
However, it will result in $O(k^2)$ bitwise operations. In other words, if we implement GF multiplication
in bit-level logic circuit, it will contain $O(k^2)$ gates. When the datapath size $k$ is large,
the area and delay of circuit will be costly.

In order to lower down the complexity of arithmetic circuit design, Massey and Omura \cite{MasseyOmura} % ref 7 in RH paper
use a new basis to represent GF elements, which is called {\bf normal basis} (NB).
A normal basis over $\Fkk$ is written in the form of
\begin{equation*}
N.B. ~~~ \N = \{\beta,\beta^2,\beta^4,\beta^8,\dots,\beta^{2^{k-1}}\}
\end{equation*}

{\bf Normal element} $\beta$ is an element from the field which is used to construct the normal basis,
and can be represent as a power of primitive element $\alpha$: 
\begin{equation*}
\beta = \alpha^t, ~~~ 1\leq t<2^k
\end{equation*}
Exponent $t$ takes value in the given range when $\N$ fulfills the definition of a basis.

Respectively, a field element in NB representation is actually
\begin{align*}
A &= (a_0,a_1,\dots,a_{k-1})_{NB} \\
  &= a_0\beta+a_1\beta^2+\cdots+a_{k-1}\beta^{2^{k-1}} \\
  &= \sum_{i=0}^{k-1} a_i\beta^{2^i}
\end{align*}

According to the definition, a normal basis is a vector where the next entry is the square of the former one.
We note that the vector is cyclic, i.e. $\beta^{2^k} = \beta$ due to {\it Fermat's little theorem}.

The addition and subtraction of elements in NB representation are similar to equation \ref{eqn:StdB}.
However, what makes NB powerful is its ease of implementation when doing multiplications and exponentiations.
The following lemmas and examples illustrate this fabulous property very well.
\begin{Lemma}[Square of NB]
\label{lem:squareNB}
In $\Fkk$, 
\begin{equation*}
(a+b)^2 = a^2 + b^2
\end{equation*}
According to the \textbf{binomial theorem}, it can be extended to
\begin{align*}
\beta^2 =&(b_0\beta + b_1\beta^2 + b_2\beta^4 + \dots + b_{k-1}\beta^{2^{k-1}})^2 \\
=& b_0^2\beta^2 + b_1^2\beta^4 + b_2^2\beta^8 + \dots + b_{k-1}^2\beta \\
=& b_{k-1}^2\beta + b_0\beta^2 + b_1\beta^4 + \dots + b_{k-2}\beta^{2^{k-1}}
\end{align*}
\end{Lemma}
This lemma concludes that the square of an element in NB equals to a simple right-cyclic shift of the bit-vector.
Obviously, StdB representation does not have this benefit.

\begin{Example}[Square of NB]
In GF $\mathbb F_{2^3}$ constructed by irreducible polynomial $x^3 + x + 1$, the standard basis is denoted as 
$\{ 1, \alpha, \alpha^2\}$ where $\alpha^3+\alpha+1=0$.
Let $\beta = \alpha^3$, then $\N = \{ \beta, \beta^2, \beta^4\}$ forms a normal basis. 
Write down element $E$ using both representations:
\begin{align*}
E &= (a_0,a_1,a_2)_{StdB} = (b_0,b_1,b_2)_{NB} \\
  &= a_0 + a_1\alpha + a_2\alpha^2 = b_0\beta + b_1\beta^2 + b_2\beta^4
\end{align*}
Compute the square of $E$ in StdB first:
\begin{align*}
E^2 &= a_0 + a_1\alpha^2 + a_2\alpha^4 \\
    &= a_0 + a_2\alpha + (a_1 + a_2)\alpha^2 \\
    &= (a_0,a_2,a_1+a_2)_{StdB}
\end{align*}
When it is computed in NB, we can make it very simple:
\begin{align*}
E^2 &= \overset{\xrightarrow{Cyclic~~shift}}{(b_0,b_1,b_2)}_{NB} \\
	&= (b_2,b_0,b_1)_{NB}
\end{align*}
\end{Example}

This example shows that convenience to use NB when computing $2^k$ power of an element.
Multiplication is a bit complicated than squaring; but when it is decomposed as bit-wise
operations, the property in lemma \ref{lem:squareNB} can be well utilized.

\begin{Example}[Bit-wise NB multiplication]
Assume there are 2 binary vectors representing 2 operands in NB over $\Fkk$: 
$A = (a_0, a_1, \dots, a_{k-1}), B = (b_0, b_1, \dots, b_{k-1})$. Note that in this example, 
by default we use normal basis representation so subscript ``NB" is skipped. Their product can also be written
as: $$C = A\times B = (c_0, c_1, \dots, c_{k-1})$$

Assume the most significant bit (MSB) of the product can be represented by a function $f_{mult}$: 
\begin{equation}
\label{eqn:multiMSB}
c_{k-1} = f_{mult}(a_0, a_1, \dots, a_{k-1}; b_0, b_1, \dots, b_{k-1})
\end{equation}
Before discussing the details of the function $f_{mult}$, we can take 
a square on both side of equation \ref{eqn:multiMSB}, i.e. $C^2 = A^2\times B^2$.
Obviously, using the property in lemma \ref{lem:squareNB}, the original second most significant bit 
becomes the new MSB because of right-cyclic shifting. 
Concretely, 
$$(c_{k-1},c_0,c_1,\dots,c_{k-2}) = (a_{k-1},a_0,a_1,\dots,a_{k-2})\times(b_{k-1},b_0,b_1,\dots,b_{k-2})$$
Note $A^2, B^2$ and $C^2$ still belong to $\Fkk$, thus as a universal function implementing MSB multiplication
over $\Fkk$, $f_{mult}$ still remains the same. As a result, the new MSB can be written as 
\begin{equation}
\label{eqn:shiftMSB}
c_{k-2} = f_{mult}(a_{k-1}, a_0, a_1, 
\dots, a_{k-2}; b_{k-1}, b_0, b_1, \dots, b_{k-2})
\end{equation}
Similarly, if we take a square again on the new equation, we can get $c_{k-3}$.
Successively we can derive all bits of product $C$ using the same function $f_{mult}$, and the only
adjustment we need to make is to right-cyclic shift 2 operands by 1 bit each time.
\end{Example}

From above example, it is known that a universal structure that implements $f_{mult}$ can be reused
$k$ times in NB multiplication over $\Fkk$. Compared to StdB, which requires a distinct design 
for every bit of multiplication, NB is less costly -- as long as we can prove $f_{mult}$ implies 
a structure with $o(k^2)$ complexity (symbol $o$ denotes ``strictly lower than bound"). 
% So our next mission is to explore the details of $f_{mult}$
% to prove it will be a relatively simple design with complexity lower than $O(k^2)$.

If we want to make the complexity of $f_{mult}$ lower than $O(k^2)$, then the best choice is to try out 
linear functions. As we know, matrix multiplication can simulate all possible combinations of 
linear functions. 
% (which is also the reason it is used as basic model to simulate the behavior of a neuron
% in neural network machine learning algorithms).
Imagine $A$ is a $k$-bit row vector and $B$ is a $k$-bit 
column vector, then the single bit product can be written as the product of matrix multiplication
$$c_{l} = A\times M\times B$$
where 
\begin{align*}
A &= (a_0,a_1,\dots,a_{k-1})\\
B &= \begin{pmatrix}
b_0 \\
b_1 \\
\vdots \\
b_{k-1}
\end{pmatrix}
\end{align*}
Moreover, $M$ is a $k\times k$ square matrix. If we can find $M$, we obtain the design of the multiplier.

\begin{Definition}[$\lambda$-Matrix]
A binary $k\times k$ matrix $M$ is used to describe the bit-wise normal basis multiplication function $f_{mult}$ where
\begin{equation}
\label{eqn:def_lambda}
c_{l} = f_{mult}(A, B) = A \times M \times B^T
\end{equation}
Symbol $B^T$ denotes vector transposition. Matrix $M$ is called $\lambda$-Matrix of
$k$-bit NB multiplication over $\Fkk$.
\end{Definition}

When taking different bits $l$ of the product in equation \ref{eqn:def_lambda}, 
we obtain a series of conjugate matrices of $M$. 
Which means instead of shifting operands $A$ and $B$, we can also shift the 
matrix.

More specifically, we denote the matrix by \emph{$l$-th $\lambda$-Matrix} as 
$$c_l = A \times M^{(l)} \cdot B^T$$
Meanwhile, the operator shifting rule in equation \ref{eqn:shiftMSB} still holds. Then we have relation 
$$c_{l-1} = A \cdot M^{(l-1)} \cdot B^T = shift(A) \cdot M^{(l)} \cdot shift(B)^T$$ which means
by right and down cyclically shifting $M^{(l-1)}$, we can get $M^{(l)}$.

\begin{Example}[NB multiplication using $\lambda$-Matrix]
Over GF $\mathbb F_{2^3}$ constructed by irreducible polynomial $\alpha^3 + \alpha + 1$, let normal element $\beta = \alpha^3$, $N = \{ \beta, \beta^2, \beta^4\}$ 
forms a normal basis. Corresponding $0$-th $\lambda$-Matrix is
\begin{equation*}
M^{(0)} = \left(
\begin{array} {lcr}
0 & 1 & 0\\
1 & 0 & 1\\
0 & 1 & 1
\end{array} \right).
\end{equation*}
i.e.,
\begin{equation*}
c_0 = (a_0\  a_1\  a_2)\left(
\begin{array} {lcr}
0 & 1 & 0\\
1 & 0 & 1\\
0 & 1 & 1
\end{array} \right)\left(
\begin{array} {lcr}
b_0\\
b_1\\
b_2
\end{array} \right)
\end{equation*}
From $0$-th $\lambda$-Matrix we can directly write down all remaining $\lambda$-Matrices:
\begin{equation*}
M^{(1)} = \left(
\begin{array} {lcr}
1 & 0 & 1\\
0 & 0 & 1\\
1 & 1 & 0
\end{array} \right)~~~~~~
M^{(2)} = \left(
\begin{array} {lcr}
0 & 1 & 1\\
1 & 1 & 0\\
1 & 0 & 0
\end{array} \right)
\end{equation*}
\end{Example}

If we generalize the definition and explore the nature of $\lambda$-Matrix, it is defined as cross-product terms from multiplication, which is 
\begin{equation}
Product~vector~C = (\sum_{i=0}^{k-1}a_i\beta^{2^i})(\sum_{j=0}^{k-1}b_j\beta^{2^j}) = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}a_ib_j\beta^{2^i}\beta^{2^j}
\end{equation}
The expressions $\beta^{2^i}\beta^{2^j}$ are referred to as cross-product terms, and can be represented by
NB, i.e.
\begin{equation}
\beta^{2^i}\beta^{2^j} = \sum_{l=0}^{k-1}\lambda_{ij}^{(l)}\beta^{2^l}, \ \ \lambda_{ij}^{(l)} \in \mathbb F_2.
\end{equation}
Substitution yields, result is an expression for l-th digit of product as showed in equation \ref{eqn:multiMSB}:
\begin{equation}
c_l = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}\lambda_{ij}^{(l)}a_ib_j
\end{equation}
$\lambda_{ij}^{(l)}$ is the entry with coordinate $(i,j)$ in $l$-th $\lambda$-Matrix.

The $\lambda$-Matrix can be implemented with XOR and AND gates in circuit design.
The very naive implementation requires $O(C_N)$ gates, where $C_N$ is the number of
nonzero entries in $\lambda$-Matrix.
There usually exists multiple NBs in $\Fkk, k>3$. If we employ a random NB, there is no mathematical
guarantee that $C_N$ has bound $o(k)$. However,
Mullin et al. \cite{mullinONB} % This citation is valid
proves that in certain GF $\mathbb F_{p^{k_{opt}}}$, there always exists at least one NB such that 
its corresponding $\lambda$-Matrix has $C_N = 2n-1$ nonzero entries. A basis with this property is
called optimal normal basis (ONB), details are introduced in appendix \ref{append:ONB}.

In practice, large size NB multipliers are usually designed in $\Fkk$ when ONB exists
to minimize the number of gates. So in the following part of this chapter and our experiments,
we only focus on ONB multipliers instead of general NB multipliers.

% After fixing this, add a whole-piece StdB vs NB example, list their cost as the conclusion
\subsection{Comparison between Standard Basis and Normal Basis}
At the end of this section, a detailed example is used to make a comparison between StdB multiplication
and NB multiplication.
\begin{Example}[Rijndael's finite field]
Rijndael uses a characteristic 2 finite field with 256 elements, which can also be called the GF $\mathbb F_{2^8}$.
Let us define the primitive element $\alpha$ using irreducible polynomial 
$\alpha^8+\alpha^7+\alpha^6+\alpha^4+\alpha^2+\alpha+1$. Coincidently, $\alpha$ is also a normal element,
i.e. $\beta = \alpha$ can construct a NB $\{\alpha,\alpha^2,\alpha^4,\alpha^8,\alpha^{16},\alpha^{32},\alpha^{64},
\alpha^{128}\}$.

We pick a pair of elements from the Rijndael's field: $A=(0100~1011)_{StdB} = (4B)_{StdB},~B=(1100~1010)_{StdB} = (CA)_{StdB}$.
First let us compute their product in StdB, the rule follows ordinary polynomial multiplication.

\begin{align*}
A\cdot B &= (\alpha^6+\alpha^3+\alpha+1)(\alpha^7+\alpha^6+\alpha^3+\alpha)\\
&= (\alpha^{13}+\alpha^{10}+\alpha^8+\alpha^7)+(\alpha^{12}+\alpha^9+\alpha^7+\alpha^6)+
	(\alpha^9+\alpha^6+\alpha^4+\alpha^3)\\
	&~~~+(\alpha^7+\alpha^4+\alpha^2+\alpha) \\
&= \alpha^{13}+\alpha^{12}+\alpha^{10}+\alpha^8+\alpha^7+\alpha^3+\alpha^2+\alpha
\end{align*}
Note that this polynomial is not the final form of the product because it needs to be
reduced modulo irreducible polynomial $\alpha^8+\alpha^7+\alpha^6+\alpha^4+\alpha^2+\alpha+1$.
This can be done using base-2 long division. Note the dividend and divisor are written in pseudo Boolean
vectors, not real Boolean vectors in any kind of bases.

% The folowing part is customized latex code mimicing binary long division
% Note \divrule coordinates may not reflect the intuitive column # in tabular. Adjust by urself!!!

\vspace{1cm}

\newdimen\digitwidth
\settowidth\digitwidth{0}
\def~{\hspace{\digitwidth}}

\def\divrule#1#2{%
\noalign{\moveright#1\digitwidth%
\vbox{\hrule width#2\digitwidth}}}
111010111\,\begin{tabular}[b]{@{}r@{}}
101001 \\ \hline
\big)\begin{tabular}[t]{@{}l@{}}
11010110001110 \\
111010111 \\ \divrule{0}{14}
~~111101101 \\
~~111010111 \\ \divrule{2}{12}
~~~~~111010110 \\
~~~~~111010111 \\ \divrule{5}{9}
~~~~~~~~~~~~~1
\end{tabular}
\end{tabular}
\vspace{0.5cm}

The final remainder is $1$, i.e. the product equals to 1 in StdB.

On the other hand, operands $A$ and $B$ can be written in NB as
$$A = (0010~1001)_{NB},~~B = (0100~0010)_{NB}$$
The $\lambda$-Matrix for $\mathbb F_{2}[x] \pmod{x^8+x^7+x^6+x^4+x^2+x+1}$
is
(Computation of $\lambda$-Matrix refers to appendix \ref{append:NB})
\begin{equation*}
M^{(0)} = \left(\begin{array}{lccccccr}
0 &0 &0 &0 &1 &0 &1 &1 \\
0 &0 &1 &1 &1 &1 &0 &0 \\
0 &1 &0 &0 &0 &0 &1 &0 \\
0 &1 &0 &0 &1 &1 &0 &1 \\
1 &1 &0 &1 &0 &1 &0 &0 \\
0 &1 &0 &1 &1 &0 &0 &1 \\
1 &0 &1 &0 &0 &0 &0 &0 \\
1 &0 &0 &1 &0 &1 &0 &1
\end{array}\right)
\end{equation*}
Taking matrix multiplication $c_0 = A\times M^{(0)}\times B^T$,
the result is $c_0 = 1$. Then by cyclic shifting $A$ and $B$
(or shifting $M^{(0)}$, either is applicable), we can successively
obtain other bits of product. The final answer is
$$C = (0000~0001)_{NB}$$
It is equivalent to the result in StdB.
\end{Example}

% From the intuition of humans, StdB multiplication is straightforward and easier to understand
% while NB is difficult to comprehend. However, if we implement both multiplications to 
% hardware multipliers, it will be clear which side a circuit designer prefers.

% May consider providing details?

Mastrovito multiplier \cite{mastro:1989} and Montgomery multiplier \cite{PT:1985} are 2 common designs
of GF multipliers using StdB. As a naive implementation of GF multiplication,
Mastrovito multiplier uses most number of gates:
$k^2$ AND gates and less than $k^2$ XOR gates \cite{Mastrovito}. Montgomery multiplier 
applies lazy reduction techniques and results in a better latency performance, while the number of gates are about
the same with Mastrovito multiplier:
$k^2$ AND gates and $k^2-k/2$ XOR gates \cite{wu:2002}.

For an 8-bit ($\mathbb F_{2^8}$) multiplier, typical design of Mastrovito multiplier consists of 218 logic gates, while 
Montgomery multiplier needs 198 gates. However, the NB multiplier reuses the $\lambda$-Matrix 
logic, so this component will only need to be implemented for once. 
Consider the definition of matrix multiplication, it needs $C_N$ AND gates to apply 
bit-wise multiplication and $C_N-1$ XOR gates to sum the intermediate products up. The number of nonzero entries
in the $\lambda$-Matrix can be counted: $C_N = 27$.
As a result, the most naive NB multiplier design (or Massey-Omura multiplier \cite{MasseyOmura})
contains 53 gates in total, which is a great saving in area cost comparing to StdB multipliers.

\section{Design a Normal Basis Multiplier on Gate Level}
The NB multiplier design consumes much less gates than ordinary StdB multiplier design, even if 
we use the most naive design. However, the modern NB multiplier design has been improved a lot from the 
very first design model proposed by Massey and Omura in 1986 \cite{MasseyOmura}. In order to 
test our approach on practical contemporary circuits, it is necessary to learn the mechanism and design 
routine of several kinds of modern NB multipliers.
\subsection{Sequential Multiplier with Parallel Outputs}
The major benefit of NB multiplier origins from the sequential design. A straightforward design implementing 
the cyclic-shift of operands and $\lambda$-Matrix logic component is the Massey-Omura multiplier.

\begin{figure}[H]
\centering{
\includegraphics[width=\textwidth]{newfig/MasseyOmura.pdf}
\caption{A typical SMSO structure of Massey-Omura multiplier}
\label{fig:MasseyOmura}}
\end{figure}

Figure \ref{fig:MasseyOmura} shows the basic architecture of a Massey-Omura multiplier. The operands
$A$ and $B$ are 2 arrays of flip-flops which allow 1-bit right-cyclic shift every clock cycle.
The logic gates in the boxes implements the matrix multiplication with $\lambda$-Matrix $M^{(0)}$, 
while each AND gate corresponds to term $a_ib_j$ and each XOR gate corresponds to addition $a_ib_j+a_{i'}b_{j'}$. 
The XOR layer has only 1 output, giving out 1 bit of product $C$ every clock cycle.

The behavior of Massey-Omura multiplier can be concluded as: pre-load operands $A,B$ and reset $C$ to 0, after 
executing for $k$ clock cycles, the data stored in flip-flop array $C$ is the product $A\times B$.
We note that there is only one output giving 1 bit of the product each clock cycle, which matches the 
definition of serial output to communication channel. Therefore this type of design is named as 
sequential multiplier with serial output (SMSO).
The SMSO architecture need $C_N$ AND gates and $C_N-1$ XOR gates, which equals to $2k-1$ AND gates and $2k-2$ XOR gates
if it is designed using ONB. In fact, the number of gates can be reduced if the multiplication is 
implemented using a conjugate of SMSO.
